---
title: "Creating visual predictive checks in R"
author: "Ron Keizer"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Creating visual predictive checks in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

# output: rmarkdown::html_vignette

The VPC is a widely used diagnostic tool in pharmacometrics (see e.g. [here](http://page-meeting.org/default.asp?abstract=1434)), most commonly created using [PsN](http://psn.sourceforge.net) and [Xpose](http://xpose.sourceforge.net), using NONMEM as the simulation engine. The aim of the current library is to provide an improved tool that is:

- a single-step process for creating a VPC. 
  - allows changing of vpc parameters such as binning / stratification upon creation of the plot, not in a separate pre-processing step.
  - easier debugging than PsN+Xpose, all data parsing and plotting in one R function
- more flexible regarding input (use simulated data from R, NONMEM, Monolix, Stan, or any other simulation tool)
- easier to customize, e.g. request any prediction / confidence interval or binning strategy upon plotting.
- easier to extend: the output is a ggplot object which can be themed and extended
- faster

## Installation

```{r, warning=FALSE, message=FALSE, error=FALSE}
    # library("devtools")
    # install_github("ronkeizer/vpc")
    library(vpc)
```
   
## How to use

- There are three main functions:
  - `vpc`: VPC for continuous data
  - `vpc_cens`: VPC for censored continuous data (e.g. data < LOQ)
  - `vpc_tte`: VPC for (repeated) time-to-event data
- The main arguments to these three function are the `sim` and `obs` arguments, which specify a simulation dataset and an observation dataset. All other arguments can be used to customize data parsing and visual appearance of the VPC such as stratification and binning.
- All three functions will return an object with parsed data and the ggplot2 object. By default, the VPC will also be plotted.

## Examples

Load the library and get the observation data and simulation data. 
In the first example, we'll use a dataset that's available in R by default (Theophylline) and also generate the simulation dataset in R. 

```{r, warning=FALSE, message=FALSE, error=FALSE}
    library(dplyr)
    library(vpc)

    ## Load the theophylline PK dataset
    obs <- Theoph
    colnames(obs) <- c("id", "wt", "dose", "time", "dv")
    obs <- obs %>%
            group_by(id) %>%  
            mutate(sex = round(runif(1))) # randomly assign a "sex" covariate
    sim <- sim_data(obs, # the design of the dataset
                    model = function(x) { # the model
                      pk_oral_1cmt (t = x$time, dose=x$dose * x$wt, ka = x$ka, ke = x$ke, cl = x$cl * x$wt, ruv = list(additive = 0.1))
                    }, 
                    theta = c(2.774, 0.0718, .0361),                 # parameter values
                    omega_mat = c(0.08854,                           # specified as lower triangle by default; 
                                  0.02421, 0.02241,                  # note: assumed that every theta has iiv, set to 0 if no iiv. 
                                  0.008069, 0.008639, 0.02862),      
                    par_names = c("ka", "ke", "cl"),                 # link the parameters in the model to the thetas/omegas
                    n = 500)
```

However, instead we could use observation and simulation data from NONMEM, e.g.:

```{r, warning=FALSE, message=FALSE, error=FALSE}
# obs <- read_table_nm("sdtab1")   # an output table with at least ID, TIME, DV
# sim <- read_table_nm("simtab1")  # a simulation file with at least ID, TIME, DV
```

The `read_table_nm()` function comes with the `vpc` library and is a fast way to read in output data from $TABLE. 

Next, the VPC can simply be created using:

```{r, fig.width=5, fig.height=5}
vpc (sim = sim, obs = obs)
```

Stratified by SEX:

```{r, fig.width=5, fig.height=5}
# vpc (sim = sim, obs = obs, strat = c("sex"))
```

Pred-correction, and plotting of data:

```{r, fig.width=5, fig.height=5}
#vpc (sim = sim, obs = obs, pred_corr = TRUE, plot_dv = TRUE)
```

With more explicit use of options, and saving the object:

```{r, fig.width=4, fig.height=4}
# vpc_dat <- vpc(sim = sim, 
#                obs = obs,                                   # supply simulation and observation dataframes
#                obs_dv = "dv",                               # these column names are the default,
#                obs_idv = "time",                            #   update these if different.
#                sim_dv = "sdv",
#                sim_idv = "time",
#                bins = c(0, 2, 4, 6, 8, 10, 25),             # specify bin separators manually
#                stratify = c("sex"),                         # multiple stratifications possible, just supply as vector
#                pi = c(0.05, 0.95),                          # prediction interval simulated data to show
#                ci = c(0.05, 0.95),                          # confidence intervals to show
#                pred_corr = FALSE,                           # perform prediction-correction?
#                plot_dv = TRUE,                              # plot observations?
#                facet = "wrap",                              # wrap stratifications, or as "row" or "column"
#                ylab = "Concentration", 
#                xlab = "Time (hrs)")
```

_Note: If you imported the data from NONMEM, the VPC function will automatically detect column names from NONMEM, such as ID, TIME, DV. If you simulated data in R or got the data from a different software, you will probably have to change the variable names for the dependent and independent variable, and the individual index._

### Censored data

The `vpc_cens()` function can be used to create a VPC for the probability of left- or right-censoring such as in the case of data <LLOQ or >ULOQ. There is no need to add a variable to the dataset to flag the censored cases, the function only requires the `lloq` or `uloq`. The example below artificially induces an LLOQ of 5 for the above model / dataset, and generates a VPC for the probability of left-censoring.

```{r, fig.width=4, fig.height=4}
#vpc_cens(sim, obs, lloq = 5)
```

### Time-to-event data

Similar to the VPC for continuous data, the VPC for TTE data requires simulated data. In general, there are two distinct approach to simulate survival data:

- *Hazard integration*: Integrate the hazard over time, and at *any possible* observation timepoint randomly draw a binary value based on the probability of observing the event. The disadvantage of this method is that it is slow due to the numerical solving of the ODEs. Also, a dataset with a dense design grid has to be used for simulation, i.e. one that has observation times at every possible timepoint that an event can occur for all individuals. 

- *Direct sampling*: Sample event times directly from the distribution used to model the data (e.g. Weibull, exponential, Gompertz). Advantages of this approach is that it is much faster, and it does not require a dense grid. The disadvantage with this approach is however that the hazard is assumed constant over time, so models with time-dependent hazards cannot easily be simulated with this approach. This approach is straightforward in R but cannot easily be implemented in NONMEM. Example will follow soon.

An example for time-to-event data is shown below. The datasets are supplied with the `vpc` library.

```{r}
data(rtte_obs_nm) 
data(rtte_sim_nm) 
```

Treat RTTE as TTE, no stratification:
```{r, fig.width=4, fig.height=4}
# vpc_tte(sim = rtte_sim_nm, 
#         obs = rtte_obs_nm, 
#         rtte = FALSE, 
#        sim_dv = "dv", obs_idv = "t", sim_idv = "t")
```

Stratified for covariate and study arm:
```{r, fig.width=6, fig.height=6}
vpc_tte(sim = rtte_sim_nm, 
        obs = rtte_obs_nm, 
        stratify = c("sex","drug"), 
        rtte = FALSE, 
        sim_dv = "dv", obs_idv = "t", sim_idv = "t")
```

Stratified for event number and study arm:
```{r, fig.width=6, fig.height=6}
vpc_tte(sim = rtte_sim_nm, 
        obs = rtte_obs_nm,
        rtte = TRUE, events = c(1:3),
        stratify = c("drug"),
        sim_dv = "dv", obs_idv = "t", sim_idv = "t")
```


